# Reproducible Research: Peer Assessment 1

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

## Loading and preprocessing the data

Ensure that the required file is present and read data into obect called "activity"
```{r}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
activity <- read.csv('activity.csv')
```

Create a date.TIME column using date and interval
```{r}
TIME <- formatC(activity$interval / 100, 2, format='f')
activity$date.TIME <- as.POSIXct(paste(activity$date, TIME),
                                 format='%Y-%m-%d %H.%M',
                                 tz='GMT')
```


from the date.TIME column, create a new date-independant TIME column for comparison of activity pattern across days
```{r}
activity$TIME <- format(activity$date.TIME, format='%H:%M:%S')
activity$TIME <- as.POSIXct(activity$TIME, format='%H:%M:%S')
```

## What is mean total number of steps taken per day?

Determine total number of steps per day
```{r}
steps_per_day <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
```

Find the mean and median
```{r}
mean_spd <- mean(steps_per_day)
median_spd <- median(steps_per_day)
mean_spd
median_spd
```

Create a histogram showing the distribution of steps per day
```{r histogram}
library(ggplot2)
qplot(steps_per_day, binwidth = 1000, xlab='Steps per Day', ylab='Frequency')
```


## What is the average daily activity pattern?

Calculate the mean steps for each five minute interval, and then put it in a data frame.
```{r}
mean.steps <- tapply(activity$steps, activity$TIME, mean, na.rm=TRUE)
daily.pattern <- data.frame(TIME=as.POSIXct(names(mean.steps)),
                            mean.steps=mean.steps)
```

Plot the average number of steps for each TIME interval across all days
```{r TIMEseriesplot}
library(scales)
ggplot(daily.pattern, aes(TIME, mean.steps)) + 
    geom_line() +
    xlab('Time') +
    ylab('Average Steps Taken') +
    scale_x_datetime(labels=date_format(format='%H:%M'))
```

Use the which.max function to determine the interval during which the most steps are taken on average
```{r}
most <- which.max(daily.pattern$mean.steps)
format(daily.pattern[most,'TIME'], format='%H:%M')
```


## Imputing missing values
Count the missing values ("NA"s)
```{r}
missing <- is.na(activity$steps)
M <- table(missing)
#NUmber of NAs: 
M["TRUE"]
```

Replace NAs with average steps for a 5-minute interval
```{r}
library(Hmisc)
activity.imputed <- activity
activity.imputed$steps <- with(activity.imputed, impute(steps, mean))
```

Let's compare the mean and median steps for each day between the original data set and the imputed data set.
```{r}
steps_per_day.imputed <- tapply(activity.imputed$steps, 
                              activity.imputed$date, sum)
mean(steps_per_day)
mean(steps_per_day.imputed)
median(steps_per_day)
median(steps_per_day.imputed)

```
Note the mean and median are now equal and greater than they were before values were imputed

Redraw teh histogram wth the imputed values
```{r histogram_imputed}
qplot(steps_per_day.imputed, binwidth = 1000, xlab='Steps per Day', ylab='Frequency')
```


## Are there differences in activity patterns between weekdays and weekends?

Create a column describing dates as either weekdays or weekends
```{r}
day_type <- function(date) {
    if (weekdays(date) %in% c('Saturday', 'Sunday')) {
        return('Weekend')
    } else {
        return('Weekday')
    }
}

day_types <- sapply(activity.imputed$date.TIME, day_type)
activity.imputed$day_type <- as.factor(day_types)
```

determine average steps for weekdays and weekends, and place each in a dataframe.
```{r}
mean.steps <- tapply(activity.imputed$steps, 
                     interaction(activity.imputed$TIME,
                                 activity.imputed$day_type),
                     mean, na.rm=TRUE)
day_type.pattern <- data.frame(TIME=as.POSIXct(names(mean.steps)),
                               mean.steps=mean.steps,
                               day_type=as.factor(c(rep('weekday', 288),
                                                   rep('weekend', 288))))
```

Plot the data for weekends and weekdays on adjacent graphs
```{r TIMEseries_daytype}
ggplot(day_type.pattern, aes(TIME, mean.steps)) + 
    geom_line() +
    xlab('Time') +
    ylab('Average Steps Taken') +
    scale_x_datetime(labels=date_format(format='%H:%M')) +
    facet_grid(day_type ~ .)
```
